State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-06-11 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
edits to the discussion

Although the system is nonlinear, the parameters of the system are linear with respect to the state. This is exploited by our procedure where the parameter estimation uses a least squares (LS) method that minimises the sum of the squared errors (of a predicted state update) with each new state vector sequence estimate. To create the least squares parameter estimator we first define the $L \times n_{\theta}$ matrix
\begin{equation}
	\mathbf{q}(\mathbf{x}_t) = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}'.
\end{equation}
Now given a state estimate sequence from the unscented Rauch Tung Striebel smoother (URTSS), we can write
\begin{eqnarray}
	\mathbf x_{1} &=& \mathbf{q}(\mathbf x_0) \boldsymbol{\theta}+\xi\mathbf x_0+\mathbf e_0 \nonumber \\
	\mathbf x_{2} &=& \mathbf{q}(\mathbf x_1) \boldsymbol{\theta}+\xi\mathbf x_1+\mathbf e_1 \nonumber \\
	&\vdots& \\
	\mathbf x_{T}&=&\mathbf{q}(\mathbf x_{T-1}) \boldsymbol{\theta}+\xi\mathbf x_{T-1}+\mathbf e_{T-1}. \nonumber \\
\end{eqnarray}
This can be written in the compact form
\begin{equation}
	\mathbf Z=\mathbf X \mathbf W+\mathbf{e} 
\end{equation}
where
\begin{small}
\begin{equation*}
	\mathbf Z=\left[
	\begin{array}{cccc}
		\mathbf x_{1}\\
		\mathbf x_{2}\\
		\vdots\\
		\mathbf x_{T}
	\end{array}
	\right],\quad \mathbf X=\left[
	\begin{array}{cccc}
		\mathbf q(\mathbf x_0)& \mathbf x_{0}\\
		\mathbf q(\mathbf x_1)& \mathbf x_{1}\\
		\vdots & \vdots\\
		\mathbf q(\mathbf x_{T-1})& \mathbf x_{T-1}
	\end{array}
	\right] 
\end{equation*}
\end{small}
and
\begin{small}
\begin{equation*}
\quad \mathbf W=\left[
	\begin{array}{cc}
		\boldsymbol{\theta} \\
		\xi
	\end{array}
	\right],\quad \mathbf{e}=\left[
	\begin{array}{cccc}
		\mathbf e_0\\
		\mathbf e_1\\
		\vdots\\
		\mathbf e_{T-1}
	\end{array}
	\right].
\end{equation*}
\end{small}
Following this, the parameters, $ \mathbf W$, can be estimated by
\begin{equation}
	\mathbf W\approx(\mathbf X^\top\mathbf X)^{-1}\mathbf X^\top\mathbf Z. 
\end{equation}
% The accuracy of the approximation is linked to the time step used in the temporal discretisation of the system, where the state transition appears more linear with a finer resolution. 
% Alternatives to the UKF are the Extended Kalman Filter (EKF)~\cite{Haykin2001} and the Sequential Monte Carlo (SMC) filter~\cite{doucet2001}. The EKF approximates the state transition equation by linearising about the current state estimates (by calculating the Jacobian) at each time instance. The linearisation maintains the Gaussianality of the model. The unscented Kalman filter approximates the posterior state density by a Gaussian distribution using a minimal set of carefully chosen sigma points, while maintaining the nonlinearity in the system. This has been shown to give superior performance over the EKF in state estimation, since the EKF maintains a first order approximation where the UKF provides an approximation accurate at least to the second order. In addition, the computation of the Jacobian used in the EKF can be problematic. SMC filtering can theoretically provide an exact posterior state density for nonlinear systems. However, currently this method is not appropriate for our problem, due to high computation demands.

@@

POSSIBLE SOURCES OF THE ERROR
\begin{itemize}
	\item slope of nonlinearity interacting with tails of Gaussian basis functions
	\item loss of high frequency info from basis function decomposition
	\item distortions of the spectral content of the basis function from the nonlinearity
	\item disturbance covariance too narrow
	\item disturbance variance too low (not exciting the kernel enough)
	\item boundary conditions
	\item
\end{itemize}

\dean{FIRST DESCRIBE AN OVER ALL SUMMARY OF WHAT WE HAVE DONE AND WHAT IS NOVEL}
This paper presents a novel framework for model reduction to facilitate practical state and parameter estimation of spatio-temporal neural fields. A basis function decomposition reduces the infinite dimensional neural field (continuous space), to a finite dimensional state-space model. The level of detail of the estimated field is related to the basis functions decomposition, providing a flexible framework where a trade-off can be made between physiological detail and computational complexity. Using the state-space model, the unscented Rauch Tung Striebel Smoother was applied for state estimation. Using the sequence of state estimates, a least squares algorithm was implemented for parameter estimation.

\dean{DISCUSS RESULTS: FREQ ANALYSIS. USE THIS TO JUSTIFY KERNEL ERRORS.}
The frequency analysis and model selection tools presented in this paper specify the sensor and field basis function arrangement required to form a state-space model that is capable of capturing the dominant cortical dynamics. In addition to providing conditions to ensure the reduced model can capture the significant spectral properties of the field, these tools can be used to better design electrophysiological recording techniques avoid spatial aliasing. The effect of the basis function decomposition will restrict the frequency content of the estimated field. This is evident in when comparing figures~\ref{fig:FieldFFT} and~\ref{fig:EstimatedFieldFFT} where the estimated field is spatially band-limited. The spatial frequency content of the field is governed by disturbance covariance and the shape of the connectivity kernel. Therefore some information will be lost due to the field decomposition. The implication of this loss of information is the high spatial frequency content of the connectivity kernel. This is demonstrated in figure~\ref{fig:KernelEstimates} where the frequency content of the estimated connectivity kernel is less than the true kernel.

\dean{DISCUSS THAT EVEN THOUGH THE KERNEL ESTIMATE IS NOT EXACT, IT IS USEFUL TO KNOW THE RATIO OF EXCITATION TO INHIBITION AND THAT RELATIVE MEASURES CAN BE ACCURATELY TRACKED}
Even though the some high spatial frequency content of the connectivity kernel is lost through the basis function decomposition, the dominant dynamics are still captured with the estimated reduced model. This is illustrated in figure~\ref{fig:FieldEstimate}. In addition, the ratio of the kernel parameters follows a tight distribution as seen in figure~\ref{fig:ParametersRatio}. This indicates that the ratio of excitation to inhibition can be tracked with this approach.

\dean{NEED TO ADD SOMETHING ABOUT THE TIME CONSTANT }

\dean{DISCUSS THE IMPLICATION OF ESTIMATING INTRACORTICAL CONNECTIVITY AND WHERE WE FIT WITHIN OTHER FUNCTIONAL CONNECTIVITY RESEARCH AREAS}
To the authors' best knowledge, this is the first paper to propose an estimation framework for intracortical connectivity from electrophysiological data. The importance of the estimation framework can be illustrated by considering that the connectivity structure at this scale can not be estimated using diffusion tensor imaging due to the near isotropic diffusion of water in grey matter~\cite{Assaf2008}. In addition, other non-model-based techniques for estimating functional connectivity, such as Granger causality~\cite{Hesse2003}, the direct transfer function~\cite{Kaminski1991} and partial directed coherence~\cite{Sameshima1999}, are restricted to the spatial scales of the recording electrodes. The model-based technique proposed in this paper estimates the connectivity structure of the number of sensors. An advantage of the auto-regressive (AR) methods is that they allow for anisotropic connectivity, which is present in cortico-cortical fibres. It should be noted, however, that the AR and the proposed model-based methods are complementary, where connectivity is estimated at different spatial scales. Therefore, we envisage that by combining these methods a more realistic model can be acquired. A recent theoretical study~\cite{Jirsa2009} has demonstrated the importance of this multi-scale structure in generating the characteristic rhythms of the brain.

\dean{DISCUSS A BIG PICTURE. WHERE THIS WORK SHOULD GO AND WHERE IT SHOULD BE APPLIED IN THE FUTURE}
Generating patient specific neural field models has the potential to have significant impact on several areas of neuroscientific research and clinical neurology. Specifically, the development of a patient-specific state-space model will allow for the application of systems theoretical techniques from engineering to be applied to neuro-dynamics. For example, state tracking could be used in brain-computer interface or epileptic seizure prediction applications, where specific regions of the state-space may indicate intent of movement or imminent seizures. Recent studies using neural field models have demonstrated the theoretical implications of specific connectivity structure of the cortical columns in neural field models, where the connectivity kernel governs where a Turing-Hopf bifurcation occurs~\cite{Hutt2005}, and the types of oscillations that can be generated~\cite{Schmidt2009}. This implies that if we can estimate the connectivity structure for an individual, then we could capture patient specific neurodynamics that lead to hyper-synchronous oscillatory states such as Turing patterns. In addition, a patient specific state-space model will allow application of control engineering techniques to robustly prevent the neuro-dynamics entering pathological regions of state-space using electrical stimulation.

@@

\dean{DISCUSS HOW IT CAN LINK THEORETICAL MODELLING STUDIES TO DATA.}
This estimation framework also has theoretical implications by establishing a framework that has the potential to allow testing of hypotheses that have been generated in theoretical studies and to validate neural field models. For example, it has been hypothesised that so-called ``bump solutions'' are a possible mechanism for short term memory formation~\cite{Coombes2005}. Using the proposed model-based framework, we can reconstruct a neural field from data, and estimate parameters and check if they correspond the the theoretically derived parameters where these bump solutions can exist. In addition, the parameters and the state-space can be explored during other cognitive tasks and states of arousal such as during sleep and under anaesthesia.

\dean{DISCUSS SHORT-COMINGS AND FUTURE WORK}
In this paper, the widths of the connectivity kernel basis functions are assumed to be known. In future work, we plan to extend this work to estimate the connectivity structure width unknown kernel basis function widths. We hope to achieve this by using a sufficiently large set of connectivity basis functions within the physiological plausible limits, and allow the estimation procedure take care of the weights. In addition, we plan to show in future work how the assumption of homogeneity can be relaxed, where heterogeneity can be modelled by careful placement of field basis functions. For example, if cortical regions have a greater density of neural populations, then more basis functions can be used in these areas. In this scenario, model selection and basis function positioning can be achieved using instantaneous frequency analysis of the the neural field or alternative model selection tools based on information criteria. Future work should also be targeted to include a more realistic synaptic response kernel, with finite rise and decay times and finite propagation times into the neural field equations.

\begin{itemize}
	\item What about the transient! The need for activity to perform estimation. \dean{To Learn You Must Perturb!} 
\end{itemize}
% 
% \section{Conclusion}\label{ConclusionSection}
% This paper presented a novel model-based framework for patient-specific state and parameter estimation of neural fields. This framework enables a fundamental link between patient-specific physiologic data and theoretical neural field models. This link has the potential to lead to advances in our basic understanding of the dynamics of the brain at the macro/mesoscopic scale. Forming a greater understanding neural dynamics at this scale may lead to new and improved treatment options for diseases such as epilepsy.

>>>>

Although the system is nonlinear, the parameters of the system are linear with respect to the state. This is exploited by our procedure where the parameter estimation uses a least squares (LS) method that minimises the sum of the squared errors (of a predicted state update) with each new state vector sequence estimate. To create the least squares parameter estimator we first define the $L \times n_{\theta}$ matrix
\begin{equation}
	\mathbf{q}(\mathbf{x}_t) = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}'.
\end{equation}
Now given a state estimate sequence from the URTSS, we can write
\begin{eqnarray}
	\mathbf x_{1} &=& \mathbf{q}(\mathbf x_0) \boldsymbol{\theta}+\xi\mathbf x_0+\mathbf e_0 \nonumber \\
	\mathbf x_{2} &=& \mathbf{q}(\mathbf x_1) \boldsymbol{\theta}+\xi\mathbf x_1+\mathbf e_1 \nonumber \\
	&\vdots& \\
	\mathbf x_{T}&=&\mathbf{q}(\mathbf x_{T-1}) \boldsymbol{\theta}+\xi\mathbf x_{T-1}+\mathbf e_{T-1}. \nonumber \\
\end{eqnarray}
This can be written in the compact form
\begin{equation}
	\mathbf Z=\mathbf X \mathbf W+\mathbf{e} 
\end{equation}
where
\begin{small}
\begin{equation*}
	\mathbf Z=\left[
	\begin{array}{cccc}
		\mathbf x_{1}\\
		\mathbf x_{2}\\
		\vdots\\
		\mathbf x_{T}
	\end{array}
	\right],\quad \mathbf X=\left[
	\begin{array}{cccc}
		\mathbf q(\mathbf x_0)& \mathbf x_{0}\\
		\mathbf q(\mathbf x_1)& \mathbf x_{1}\\
		\vdots & \vdots\\
		\mathbf q(\mathbf x_{T-1})& \mathbf x_{T-1}
	\end{array}
	\right] 
\end{equation*}
\end{small}
and
\begin{small}
\begin{equation*}
\quad \mathbf W=\left[
	\begin{array}{cc}
		\boldsymbol{\theta} \\
		\xi
	\end{array}
	\right],\quad \mathbf{e}=\left[
	\begin{array}{cccc}
		\mathbf e_0\\
		\mathbf e_1\\
		\vdots\\
		\mathbf e_{T-1}
	\end{array}
	\right].
\end{equation*}
\end{small}
Following this, the parameters, $ \mathbf W$, can be estimated by
\begin{equation}
	\mathbf W\approx(\mathbf X^\top\mathbf X)^{-1}\mathbf X^\top\mathbf Z. 
\end{equation}
% The accuracy of the approximation is linked to the time step used in the temporal discretisation of the system, where the state transition appears more linear with a finer resolution. 
% Alternatives to the UKF are the Extended Kalman Filter (EKF)~\cite{Haykin2001} and the Sequential Monte Carlo (SMC) filter~\cite{doucet2001}. The EKF approximates the state transition equation by linearising about the current state estimates (by calculating the Jacobian) at each time instance. The linearisation maintains the Gaussianality of the model. The unscented Kalman filter approximates the posterior state density by a Gaussian distribution using a minimal set of carefully chosen sigma points, while maintaining the nonlinearity in the system. This has been shown to give superior performance over the EKF in state estimation, since the EKF maintains a first order approximation where the UKF provides an approximation accurate at least to the second order. In addition, the computation of the Jacobian used in the EKF can be problematic. SMC filtering can theoretically provide an exact posterior state density for nonlinear systems. However, currently this method is not appropriate for our problem, due to high computation demands.

@@

POSSIBLE SOURCES OF THE ERROR
\begin{itemize}
	\item slope of nonlinearity interacting with tails of Gaussian basis functions
	\item loss of high frequency info from basis function decomposition
	\item distortions of the spectral content of the basis function from the nonlinearity
	\item disturbance covariance too narrow
	\item disturbance variance too low (not exciting the kernel enough)
	\item boundary conditions
	\item more...
\end{itemize}

This paper presents a novel framework for model reduction to facilitate practical state and parameter estimation of spatio-temporal neural fields. A basis function decomposition reduces the infinite dimensional neural field (continuous space), to a finite dimensional state-space model. The level of detail of the estimated field is related to the basis functions decomposition, providing a flexible framework where a trade-off can be made between physiological detail and computational complexity. Using the state-space model, the unscented Rauch Tung Striebel smoother was applied for state estimation. Using the sequence of state estimates, a least squares algorithm was implemented for parameter estimation. Though our novel procedure, we have demonstrated that it is theoretically possible to estimate intracortical connectivity and the synaptic time constant.

The frequency analysis and model selection tools presented in this paper specify the sensor and field basis function arrangement required to form a state-space model that is capable of capturing the dominant cortical dynamics. In addition to providing conditions to ensure the reduced model can capture the significant spectral properties of the field, these new tools can be used to better design electrophysiological recording techniques avoid spatial aliasing. This analysis is becoming increasingly relevant as recording systems are becoming more sophisticated with higher density recordings~\cite{Brinkmann2009}. The effect of the basis function decomposition will restrict the frequency content of the estimated field. This is evident in when comparing Figures~\ref{fig:FieldFFT} and~\ref{fig:EstimatedFieldFFT} where the estimated field is spatially band-limited. The spatial frequency content of the field is governed by disturbance covariance and the shape of the connectivity kernel.
% 
% In practice Even though the some high spatial frequency content of the connectivity kernel is lost through the basis function decomposition, the dominant dynamics are still captured with the estimated reduced model. This is illustrated in figure~\ref{fig:FieldEstimate}. In addition, the ratio of the kernel parameters follows a tight distribution as seen in figure~\ref{fig:ParametersRatio}. This indicates that the ratio of excitation to inhibition can be tracked with this approach.

To the authors' best knowledge, this is the first paper to propose an estimation framework for intracortical connectivity from electrophysiological data. The importance of the estimation framework can be illustrated by considering that the connectivity structure at this scale can not be estimated using diffusion tensor imaging due to the near isotropic diffusion of water in grey matter~\cite{Assaf2008}. In addition, other non-model-based techniques for estimating functional connectivity, such as Granger causality~\cite{Hesse2003}, the direct transfer function~\cite{Kaminski1991} and partial directed coherence~\cite{Sameshima1999}, are restricted to the spatial scales of the recording electrodes. The model-based technique proposed in this paper estimates the connectivity structure of the number of sensors. An advantage of the auto-regressive (AR) methods is that they allow for anisotropic connectivity, which is present in cortico-cortical fibres. It should be noted, however, that the AR and the proposed model-based methods are complementary, where connectivity is estimated at different spatial scales. Therefore, we envisage that by combining these methods a more realistic model can be acquired. A recent theoretical study~\cite{Jirsa2009} has demonstrated the importance of this multi-scale structure in generating the characteristic rhythms of the brain.

Generating patient specific neural field models has the potential to have significant impact on several areas of neuroscientific research and clinical neurology. Specifically, the development of a patient-specific state-space model will allow for the application of systems theoretical techniques from engineering to be applied to neuro-dynamics. For example, state tracking could be used in brain-computer interface or epileptic seizure prediction applications, where specific regions of the state-space may indicate intent of movement or imminent seizures. Recent studies using neural field models have demonstrated the theoretical implications of specific connectivity structure of the cortical columns in neural field models, where the connectivity kernel governs where a Turing-Hopf bifurcation occurs~\cite{Hutt2005}, and the types of oscillations that can be generated~\cite{Schmidt2009}. This implies that if we can estimate the connectivity structure for an individual, then we could capture patient specific neurodynamics that lead to hyper-synchronous oscillatory states such as Turing patterns. In addition, a patient specific state-space model will allow application of control engineering techniques to robustly prevent the neuro-dynamics entering pathological regions of state-space using electrical stimulation.

@@

This estimation framework also has theoretical implications by establishing a framework that has the potential to allow testing of hypotheses that have been generated in theoretical studies and to validate neural field models. For example, it has been hypothesised that so-called ``bump solutions'' are a possible mechanism for short term memory formation~\cite{Coombes2005}. Using the proposed model-based framework, we can reconstruct a neural field from data, and estimate parameters and check if they correspond the the theoretically derived parameters where these bump solutions can exist. In addition, the parameters and the state-space can be explored during other cognitive tasks and states of arousal such as during sleep and under anaesthesia.

In this paper, the widths of the connectivity kernel basis functions are assumed to be known. In future work, we plan to extend this work to estimate the connectivity structure width unknown kernel basis function widths. We hope to achieve this by using a sufficiently large set of connectivity basis functions within the physiological plausible limits, and allow the estimation procedure take care of the weights. In addition, we plan to show in future work how the assumption of homogeneity can be relaxed, where heterogeneity can be modelled by careful placement of field basis functions. For example, if cortical regions have a greater density of neural populations, then more basis functions can be used in these areas. In this scenario, model selection and basis function positioning can be achieved using instantaneous frequency analysis of the the neural field or alternative model selection tools based on information criteria. Future work should also be targeted to include a more realistic synaptic response kernel, with finite rise and decay times and finite propagation times into the neural field equations.

% 
% \section{Conclusion}\label{ConclusionSection}
% This paper presented a novel model-based framework for patient-specific state and parameter estimation of neural fields. This framework enables a fundamental link between patient-specific physiologic data and theoretical neural field models. This link has the potential to lead to advances in our basic understanding of the dynamics of the brain at the macro/mesoscopic scale. Forming a greater understanding neural dynamics at this scale may lead to new and improved treatment options for diseases such as epilepsy.
